The Two Dualities of Computation: Negative and Fractional Types

### 2012-03-10 rational.tex
intro.

\paragraph*{Negative Types.} 
Consider the following algebraic manipulation relating a natural number {{a}}
to itself (ignoring the dotted line for a moment):
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire1.pdf}
}
\end{center}
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~{{a}} (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, this computation flows along the dotted
line with the following intermediate steps:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being in the
  left summand of the type {{a+0}}; we indicate this value as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}
The example illustrates that the empty type and negative types have a
fascinating computational interpretation with some flavor of continuations.
What is even more fascinating is that the above computation can be realized
with a completely different concurrent schedule as
follows. \emph{Simultaneously} the execution begins at line~(0) with the
value {{20}}, at line (2) with the value {{right (right 20)}} flowing forwards
and the value {{right (left 20)}} flowing backwards, and at line (3) with the
value {{left (left 20)}} flowing forwards and the value {{left (right 20)}}
flowing backwards. We now have three independent subcomputations proceeding in
parallel: the positive value created at line (2) which can flow to the
consumer; the negative value created at line (3) which can pay for the debt
created at line (2); and the initial value from line (0) which can flow to
the required forward-flowing value needed at line (3). At some intuitive
level, one might understand the situation as modeling a credit card payment
in which money is created at line (2) for the consumer together with a
corresponding debt. The producer's bank pays for the debt at line (3) and
waits for the actual money to arrive from the producer.

There are at least three fundamental points about this example that must be
emphasized:
\begin{itemize}
\item As the example illustrates, negative types corresponds to ``debts.''
  It would clearly be disastrous if debts could be deleted or
  duplicated. This simple observation explains why negative types are much
  simpler and much more appealing in a framework where information is
  guaranteed to be preserved. In previous work that used negative types (see
  Sec.~\ref{sec:related}), complicated mechanisms are typically needed to
  constrain the propagation and use of negative values because the
  surrounding computational framework is, generally speaking, careless in its
  treatment of information.
\item The appeal of negative types can already be gleaned from the credit
  card analogy. The main reason credit card transactions are convenient is
  because they disentangle the propagation of the resources (money) from the
  propagation of the services. Not every transaction needs both the resources
  and services to be brought together: it is sufficient to have a promise
  that the demand for resources will be somehow satisfied, as long as the
  infrastructure can be trusted with such promises. Computationally, this
  means that negative types allow us to break the dependency between
  producers and consumers and let each compute independently.
\item Consider a value of type {{-a + b}}. The above discussion suggests that
  this value demands {{a}} and produces {{b}}, i.e., that it is a
  higher-order funtion mapping {{a}} to {{b}}. Dually and as the underlying
  categorical semantics justifies (see Sec.~\ref{sec:neg}), the same value is
  a continuation which produces {{b}} and then demands {{a}}, which
  introduces a twist on how the function can be used: get the goods {{b}} now
  and pay {{a}} later.
\end{itemize}

\paragraph*{Fractional Types.} 
Consider the following realization of boolean negation written using a
pattern-matching style:
\begin{verbatim}
notB b = case b of 
  False -> True
  True -> False
\end{verbatim}
In a functional language, we are not used to treating the pattern-matching
clauses \verb|False -> True| and \verb|True -> False| as first-class entities
but fractional types allow us to do just that! Informally, we can treat each
of the pattern-matching clauses as a first-class value of type 
{{(1/bool) * bool}}. 
A small but complete example will illustrate the main aspects of this idea.

>>>>

\paragraph*{Negative Types.} 
Consider the following algebraic manipulation relating a natural number {{a}}
to itself (ignoring the dotted line for a moment):
\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire1.pdf}
}
\end{center}
Although seemingly pointless, this algebraic proof corresponds, in our model,
to an isomorphism of type {{a <-> a}} with a non-trivial and interesting
computational interpretation. The witness for this isomorphism is a
computation that takes a value of type~{{a}} (say \$20.00) and eventually
produces the same \$20.00 value as its output. As the semantics of
Sec.~\ref{sec:rat} formalizes, this computation `flows along the dotted
line' with the following intermediate steps:
\begin{itemize}
\item We start at line (0) with \$20.00; 
\item We proceed to line (1) with the same \$20.00 but tagged as being
  in the left summand of the sum type `{{a+0}}'; we indicate this value
  as {{left 20}}
\item We continue to line (2) with the same value {{left 20}};
\item At line (3), as a result of re-association the tag on the \$20.00
  changes to indicate that it is in the left-left summand, i.e., the value is
  now {{left (left 20)}};
\item At line (4), we find ourselves needing to produce a value of type 0
  which is impossible; this signals the beginning of a reverse execution
  which sends us back to line (3) with a value {{left (right 20)}};
\item Execution continues in reverse to line (2) with the value 
  {{right (left 20)}};
\item At line (1) we find ourselves again facing an empty type so we reverse
  execution again; we go to line (2) with a value {{right (right 20)}};
\item We proceed to line (3) and (4) with the value {{right 20}};
\item We finally reach line 5 with the value {{20}}.
\end{itemize}
The example illustrates that the empty type and negative types have a
fascinating computational interpretation as values that backtrack to
satisfy dependancies and have some flavor of continuations.

\paragraph*{Fractional Types.} What is even more fascinating is that
the above computation can be realized with a completely different
concurrent schedule as follows, with fractional types.

\begin{center}
\scalebox{1.1}{
\includegraphics{diagrams/thesis/algebra-wire2.pdf}
}
\end{center}

%% The execution begins at line~(0) with the value {{20}}, at line (2)
%% with the value {{right (right 20)}} flowing forwards and the value
%% {{right (left 20)}} flowing backwards, and at line (3) with the value
%% {{left (left 20)}} flowing forwards and the value {{left (right 20)}}
%% flowing backwards.

%\emph{Simultaneously} 
The execution begins at line~(0) with the value {{20}}. At line~(1)
values {{20}} and {{()}} flow forward, corresponding to the type
`{{a*1}}'. At line~(2) three independent subcomputations \emph{proceed
  concurrently}: the intial value {{20}} flowing from line (1) and two
entangled values, {{1/x}} and {{x}}, such that specifying one
implicitly specifies the dependant other. These three subcomputations
exist \emph{at the same time}. At line (3), the dependant {{1/x}} is
fixed to be the value {{20}}, thereby making fixing values of all
three branches.
%% the positive value created at line (2) which can flow to the
%% consumer; the negative value created at line (3) which can pay for the
%% debt created at line (2); and the initial value from line (0) which
%% can flow to the required forward-flowing value needed at line (3). 
At some intuitive level, one might understand the situation as
modeling a credit card payment in which money is created at line (2)
for the consumer together with a corresponding debt. The producer's
bank pays for the debt at line (3) and waits for the actual money to
arrive from the producer.

There are at least two fundamental points about this example that must be
emphasized:
\begin{itemize}
\item As the example illustrates, negative types corresponds to
  ``debts'' satified by backtracking and fractional types correspond
  to debts satisfied by constraint propagation.  It would clearly be
  disastrous if debts could be deleted or duplicated. This simple
  observation explains why these types are much simpler and much more
  appealing in a framework where information is guaranteed to be
  preserved. In previous work that used negative types (see
  Sec.~\ref{sec:related}), complicated mechanisms are typically needed
  to constrain the propagation and use of negative values because the
  surrounding computational framework is, generally speaking, careless
  in its treatment of information.
\item The appeal of negative and fractional types can already be
  gleaned from the credit card analogy. The main reason credit card
  transactions are convenient is because they disentangle the
  propagation of the resources (money) from the propagation of the
  services. Not every transaction needs both the resources and
  services to be brought together: it is sufficient to have a promise
  that the demand for resources will be somehow satisfied, as long as
  the infrastructure can be trusted with such
  promises. 
%% Computationally, this means that negative types allow us
%%   to break the dependency between producers and consumers and let each
%%   compute independently.

%% \item Consider a value of type {{-a + b}}. The above discussion suggests that
%%   this value demands {{a}} and produces {{b}}, i.e., that it is a
%%   higher-order funtion mapping {{a}} to {{b}}. Dually and as the underlying
%%   categorical semantics justifies (see Sec.~\ref{sec:neg}), the same value is
%%   a continuation which produces {{b}} and then demands {{a}}, which
%%   introduces a twist on how the function can be used: get the goods {{b}} now
%%   and pay {{a}} later.
\end{itemize}

\paragraph*{Functions and Dependancy Splitting.} 
Consider the following realization of boolean negation written using a
pattern-matching style:
\begin{verbatim}
notB b = case b of 
  False -> True
  True -> False
\end{verbatim}
In a functional language, we are not used to treating the pattern-matching
clauses \verb|False -> True| and \verb|True -> False| as first-class entities
but fractional types allow us to do just that! Informally, we can treat each
of the pattern-matching clauses as a first-class value of type 
{{(1/bool) * bool}}. 
A small but complete example will illustrate the main aspects of this idea.
